---
title: "Quarto Computations"
format:
  html:
    df-print: paged
    code-fold: true
    code-tools: true
    code-link: true
---

This dataset contains a subset of the fuel economy data from the EPA. Specifically, we use the `mpg` dataset from the **ggplot2** package.

```{r}
#| label: load-packages
#| echo: false

library(ggplot2)
```

```{r}
mpg
```


The visualization below shows a positive, strong, and linear relationship between the city and highway mileage of these cars. Additionally, mileage is higher for cars with fewer cylinders.

```{r}
#| label: scatterplot

ggplot(mpg, aes(x = hwy, y = cty, color = cyl)) +
  geom_point(alpha = 0.5, size = 2) +
  scale_color_viridis_c() +
  theme_minimal()
```

```{r}
#| label: fig-scatterplot
#| fig-cap: "City and highway mileage for 38 popular models of cars."
#| fig-alt: "Scatterplot of city vs. highway mileage for cars, where points are colored by the number of cylinders. The plot displays a positive, linear, and strong relationship between city and highway mileage, and mileage increases as the number cylinders decreases."
#| fig-width: 6
#| fig-height: 3.5

ggplot(mpg, aes(x = hwy, y = cty, color = cyl)) +
  geom_point(alpha = 0.5, size = 2) +
  scale_color_viridis_c() +
  theme_minimal()
```

@fig-scatterplot shows a positive, strong, and linear relationship between the city and highway mileage of these cars.

```{r}
#| label: fig-mpg
#| fig-cap: "City and highway mileage for 38 popular models of cars."
#| fig-subcap:
#|   - "Color by number of cylinders"
#|   - "Color by engine displacement, in liters"
#| layout-ncol: 2
#| column: page

ggplot(mpg, aes(x = hwy, y = cty, color = cyl)) +
  geom_point(alpha = 0.5, size = 2) +
  scale_color_viridis_c() +
  theme_minimal()

ggplot(mpg, aes(x = hwy, y = cty, color = displ)) +
  geom_point(alpha = 0.5, size = 2) +
  scale_color_viridis_c(option = "E") +
  theme_minimal()
```

The plots in @fig-mpg show the relationship between city and highway mileage for 38 popular models of cars.
In @fig-mpg-1 the points are colored by the number of cylinders while in @fig-mpg-2 the points are colored by engine displacement.


